A note on the instability and degeneracy of deep learning models

### 2016-11-29 writing/draft.Rmd
small updates from meeting

### Near-degeneracy
In Random Graph Model theory, *model degeneracy* means there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model [@handcock2003assessing]. For random graph models, $\mathcal{X}$ denotes all possible graphs that can be constructed from a set of nodes and an exponentially parameterized random graph model has a distribution of the form 
\begin{align*}
f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}, \boldsymbol x \in \mathcal{X},
\end{align*}
where $\boldsymbol \theta \in \Theta \subset \mathbb{R}^q$ is the model parameter, and $\boldsymbol t: \mathcal{X} \rightarrow \mathbb{R}^q$ is a vector of statistics based on the adjacency matrix of a graph. Here, as earlier, $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)$ is the normalizing function. Let $C$ denote the convex hull of the potential outcomes of sufficient statistics, $\{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$. @handcock2003assessing classify an exponentially parametrized random graph model at $\boldsymbol \theta$ as *near-degenerate* if the mean value of the vector of sufficient statistics under $\boldsymbol \theta$, $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$, is close to the boundary of $C$. This makes sense because if a model is near-degenerate in the sense that only a small number of elements of the sample space $\mathcal{X}$ have positive probability, the expected value $\text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$ is an average of that same small number of values of $t( \boldsymbol x)$ and can be expected to not be pulled deep into the interior of $C$.

@@

In other words, if the probability ratio (\ref{eqn:elpr}) is too large, then a RBM model sequence will exhibit large probability shifts for very small changes in data configurations (i.e., will exhibit instability).

@@

    \begin{align*}
    \gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^\nv\sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right) \text{ and } C_{2} < C_{1}
    \end{align*}
    This is the method of @li2014biclustering. We will refer to this method as Bayes with Trick Prior and Latent Variables (BwTPLV).
2. *A truncated Normal prior.* Here we use independent spherical normal distributions as priors for $\boldsymbol \theta_{main}$ and $\boldsymbol \theta_{interaction}$ with $\sigma_{interaction} < \sigma_{main}$, truncated at $3\sigma_{main}$ and $3\sigma_{interaction}$, respectively. Full conditional distributions are not conjugate, and simulation from the posterior was accomplished using a geometric adaptive Metropolis Hastings step [@zhou2014some] and calculation of likelihood normalizing constant. (This computation is barely feasible for a problem of this size and would be infeasible for larger problems.) Here the $h_j$ are carried along in the MCMC implementation as latent variables. We will refer to this method as Bayes with Truncated Normal prior and Latent Variables (BwTNLV).
3. *A truncated Normal prior and marginalized likelihood.* Here we marginalize out $\boldsymbol h$ in $f_{\boldsymbol \theta}(\boldsymbol x)$, and use the truncated Normal priors applied to the marginal probabilities for visible variables given by
    \begin{align*}
    g_{\boldsymbol \theta}(\boldsymbol v) \propto \sum\limits_{\boldsymbol h \in \mathcal{C}^\nh} \exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right), \boldsymbol v \in \mathcal{C}^\nv.
    \end{align*}
    We will refer to this method as Bayes with Truncated Normal prior and Marginalized Likelihood (BwTNML).

>>>>

### Near-degeneracy
In Random Graph Model theory, *model degeneracy* means there is a disproportionate amount of probability placed on only a few elements of the sample space, $\mathcal{X}$, by the model [@handcock2003assessing]. For random graph models, $\mathcal{X}$ denotes all possible graphs that can be constructed from a set of nodes and an exponentially parameterized random graph model has a distribution of the form 
\begin{align*}
f_{\boldsymbol \theta} (\boldsymbol x) = \frac{\exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)}{\gamma(\boldsymbol \theta)}, \boldsymbol x \in \mathcal{X},
\end{align*}
where $\boldsymbol \theta \in \Theta \subset \mathbb{R}^q$ is the model parameter, and $\boldsymbol t: \mathcal{X} \rightarrow \mathbb{R}^q$ is a vector of statistics based on the adjacency matrix of a graph. Here, as earlier, $\gamma(\boldsymbol \theta) = \sum_{\boldsymbol x \in \mathcal{X}} \exp\left(\boldsymbol \theta^T \boldsymbol t(\boldsymbol x)\right)$ is the normalizing function. Let $C$ denote the convex hull of the potential outcomes of sufficient statistics, $\{\boldsymbol t(\boldsymbol x): \boldsymbol x \in \mathcal{X}\}$. @handcock2003assessing classify an exponentially parametrized random graph model at $\boldsymbol \theta$ as *near-degenerate* if the mean value of the vector of sufficient statistics under $\boldsymbol \theta$, $\boldsymbol \mu(\boldsymbol \theta) = \text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$, is close to the boundary of $C$. This makes sense because if a model is near-degenerate in the sense that only a small number of elements of the sample space $\mathcal{X}$ have positive probability, the expected value $\text{E}_{\boldsymbol \theta}\boldsymbol t( \boldsymbol X)$ is an average of that same small number of values of $t( \boldsymbol x)$ and can be expected to *not* be pulled deep into the interior of $C$.

@@

In other words, if the probability ratio (\ref{eqn:elpr}) is too large, then a RBM model sequence will exhibit large probability shifts for very small changes in data configurations (i.e., will exhibit instability). Recall the applied example of RBM models as a means to classify images. This instability result manifests itself as a one pixel change in an image (one component of the visible vector) resulting in a large shift in the probability, which in turn could result in a vastly different classification of the image. Examples of this behavior have been presented in @szegedy2013intriguing for other deep learning models, in which a one pixel change in a test image results in a wildly different classification.

@@

    \begin{align*}
    \gamma(\boldsymbol \theta) = \sum\limits_{\boldsymbol x \in \mathcal{C}^{\nh + \nv}}\exp\left(\sum\limits_{i = 1}^\nv\sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^V\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right) \text{ and } C_{2} < C_{1}
    \end{align*}
    This is the method of @li2014biclustering. We will refer to this method as Bayes with Trick Prior and Latent Variables (BwTPLV).
2. *A truncated Normal prior.* Here we use independent spherical normal distributions as priors for $\boldsymbol \theta_{main}$ and $\boldsymbol \theta_{interaction}$ with $\sigma_{interaction} < \sigma_{main}$, *truncated* at $3\sigma_{main}$ and $3\sigma_{interaction}$, respectively. Full conditional distributions are not conjugate, and simulation from the posterior was accomplished using a geometric adaptive Metropolis Hastings step [@zhou2014some] and calculation of likelihood normalizing constant. (This computation is barely feasible for a problem of this size and would be infeasible for larger problems.) Here the $h_j$ are carried along in the MCMC implementation as latent variables. We will refer to this method as Bayes with Truncated Normal prior and Latent Variables (BwTNLV).
3. *A truncated Normal prior and marginalized likelihood.* Here we marginalize out $\boldsymbol h$ in $f_{\boldsymbol \theta}(\boldsymbol x)$, and use the truncated Normal priors applied to the marginal probabilities for visible variables given by
    \begin{align*}
    g_{\boldsymbol \theta}(\boldsymbol v) \propto \sum\limits_{\boldsymbol h \in \mathcal{C}^\nh} \exp\left(\sum\limits_{i = 1}^\nv \sum\limits_{j=1}^\nh \theta_{ij} v_i h_j + \sum\limits_{i = 1}^\nv\theta_{v_i} v_i + \sum\limits_{j = 1}^\nh\theta_{h_j} h_j\right), \boldsymbol v \in \mathcal{C}^\nv.
    \end{align*}
    We will refer to this method as Bayes with Truncated Normal prior and Marginalized Likelihood (BwTNML).

### 2016-11-29 writing/note.Rmd
small updates from meeting

For $\boldsymbol X = X_1, \dots, X_N$ discrete random variables with sample space $\mathcal{X}^N$, $|\mathcal{X}| < \infty$, consider an exponential family model for $\boldsymbol X$ with probability mass function of the form
$$
p_{N, \boldsymbol \lambda}(\boldsymbol x) = \exp\left[\boldsymbol\eta^T(\boldsymbol \lambda) \boldsymbol g_N(\boldsymbol x) - \psi(\boldsymbol \lambda)\right], \qquad \boldsymbol x \in \mathcal{X}^N,
$$ 
for fixed positive integers $k$ and $L$ denoting the dimensions of the parameter and natural parameters spaces respectively, $\boldsymbol \lambda \in \Lambda \subset \mathbb{R}^{k}$, $\boldsymbol \eta : \mathbb{R}^k \mapsto \mathbb{R}^L$, $\boldsymbol g_N : \mathcal{X}^N \mapsto \mathbb{R}^L$ a vector of sufficient statistics, 
$$
\psi(\boldsymbol \lambda) = \log \sum\limits_{\boldsymbol x \in \mathcal{X}^N}\exp\left[\boldsymbol \eta^T(\boldsymbol \lambda) \boldsymbol g_N(\boldsymbol x) \right], \qquad \boldsymbol \lambda \in \Lambda,
$$
the normalizing function, and $\Lambda = \{\boldsymbol \lambda \in \mathbb{R}^k : \psi(\boldsymbol \lambda) < \infty, k \le q(N) \}$ is the parameter space.

Defining $\boldsymbol \theta_N=\lambda$ to a sequence of elements of $\Lambda \subset \mathbb{R}^k$ and noting that $p_{N, \boldsymbol \lambda}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, these discrete exponential family models are special cases of the FSFS models. Such exponential models arise with spatial data on a lattice [@besag1974spatial], network data [@wasserman1994social; @handcock2003assessing], and even binomial sampling with $N$ iid Bernoulli random variables. For these exponential models, the dimension of the parameter $\boldsymbol \theta_N$ is the same for each $N$ (i.e., $\boldsymbol \theta_N$ lies in a parameter space of fixed Euclidean dimension $k$).  This need not be true for other types of FSFS models considered in Sections \ref{restricted-boltzmann-machines}-\ref{deep-learning}. @schweinberger2011instability considered instability in such exponential models (e.g., for random graphs) for sequences of fixed parameters $\theta_N=\lambda\in\mathbb{R}^k$, $N \geq 1$, of non-varying dimension $k$.

@@

\begin{remark}
There is a further generalization the notion of instability in Definition \ref{defn:instab} meant to address independent replications of data sequences. That is, one might consider data as $n$ independent and identically distributed replications $\boldsymbol X_1, \dots, \boldsymbol X_n$, where each $\boldsymbol X_i=(X_{i,1},\ldots,X_{i,N}) \in \mathcal{X}^N$ follows a common FSFS model with probabilities $P_{\theta_N}(\boldsymbol x)>0$,  $\boldsymbol x\in\mathcal{X}^N$ and $|\mathcal{X}|<\infty$, for $i=1,\ldots,n$.  This leads to a total of $n*N$ random variables in the joint model. However, the definition of S-unstable and Propositions \ref{prop:instab}-\ref{prop:degen} still hold for such iid replications. This is because $\left(\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n$ is the largest probability possible under the joint model for the $n$ replications while $\left(\min\limits_{\boldsymbol x \in \mathcal{X}}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n$ is the smallest probability. Thus, 
$$
\frac{\text{extremal log-probability ratio}}{\text{\# random variables in the model}} \equiv \frac{1}{n*N}\log \left[\frac{\left(\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n}{\left(\min\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta}(\boldsymbol x)\right)^n}\right] = \frac{1}{N} \log \left[\frac{\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)}{\min\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta}(\boldsymbol x)}\right] = \frac{1}{N} \text{ELPR}(\boldsymbol \theta_N),
$$
implying that the definition of an S-unstable FSFS model sequence is invariant to the level ($n$) of independent replication. Consequently, overall model instabilities may be characterized by those of one observation from the common FSFS model.   
\end{remark}

>>>>

For $\boldsymbol X = X_1, \dots, X_N$ discrete random variables with sample space $\mathcal{X}^N$, $|\mathcal{X}| < \infty$, consider an exponential family model for $\boldsymbol X$ with probability mass function of the form
$$
p_{\boldsymbol \lambda}(\boldsymbol x) = \exp\left[\boldsymbol\eta^T(\boldsymbol \lambda) \boldsymbol g_N(\boldsymbol x) - \psi(\boldsymbol \lambda)\right], \qquad \boldsymbol x \in \mathcal{X}^N,
$$ 
for fixed positive integers $k$ and $L$ denoting the dimensions of the parameter and natural parameters spaces respectively, $\boldsymbol \lambda \in \Lambda \subset \mathbb{R}^{k}$, $\boldsymbol \eta : \mathbb{R}^k \mapsto \mathbb{R}^L$, $\boldsymbol g_N : \mathcal{X}^N \mapsto \mathbb{R}^L$ a vector of sufficient statistics, 
$$
\psi(\boldsymbol \lambda) = \log \sum\limits_{\boldsymbol x \in \mathcal{X}^N}\exp\left[\boldsymbol \eta^T(\boldsymbol \lambda) \boldsymbol g_N(\boldsymbol x) \right], \qquad \boldsymbol \lambda \in \Lambda,
$$
the normalizing function, and $\Lambda = \{\boldsymbol \lambda \in \mathbb{R}^k : \psi(\boldsymbol \lambda) < \infty, k \le q(N) \}$ is the parameter space.

Defining $\boldsymbol \theta_N=\lambda$ to a sequence of elements of $\Lambda \subset \mathbb{R}^k$ and noting that $p_{\boldsymbol \lambda}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, these discrete exponential family models are special cases of the FSFS models. Such exponential models arise with spatial data on a lattice [@besag1974spatial], network data [@wasserman1994social; @handcock2003assessing], and even binomial sampling with $N$ iid Bernoulli random variables. For these exponential models, the dimension of the parameter $\boldsymbol \theta_N$ is the same for each $N$ (i.e., $\boldsymbol \theta_N$ lies in a parameter space of fixed Euclidean dimension $k$).  This need not be true for other types of FSFS models considered in Sections \ref{restricted-boltzmann-machines}-\ref{deep-learning}. @schweinberger2011instability considered instability in such exponential models (e.g., for random graphs) for sequences of fixed parameters $\theta_N=\lambda\in\mathbb{R}^k$, $N \geq 1$, of non-varying dimension $k$.

@@

\begin{remark}
There is a further generalization the notion of instability in Definition \ref{defn:instab} meant to address independent replications of data sequences. That is, one might consider data as $n$ independent and identically distributed replications $\boldsymbol X_1, \dots, \boldsymbol X_n$, where each $\boldsymbol X_i=(X_{i,1},\ldots,X_{i,N}) \in \mathcal{X}^N$ follows a common FSFS model with probabilities $P_{\theta_N}(\boldsymbol x)>0$,  $\boldsymbol x\in\mathcal{X}^N$ and $|\mathcal{X}|<\infty$, for $i=1,\ldots,n$.  This leads to a total of $n*N$ random variables in the joint model. However, the definition of S-unstable and Propositions \ref{prop:instab}-\ref{prop:degen} still hold for such iid replications. This is because $\left(\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n$ is the largest probability possible under the joint model for the $n$ replications while $\left(\min\limits_{\boldsymbol x \in \mathcal{X}}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n$ is the smallest probability. Thus, 
$$
\frac{\text{extremal log-probability ratio}}{\text{\# random variables in the model}} \equiv \frac{1}{n*N}\log \left[\frac{\left(\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n}{\left(\min\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)\right)^n}\right] = \frac{1}{N} \log \left[\frac{\max\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)}{\min\limits_{\boldsymbol x \in \mathcal{X}^N}P_{\boldsymbol \theta_N}(\boldsymbol x)}\right] = \frac{1}{N} \text{ELPR}(\boldsymbol \theta_N),
$$
implying that the definition of an S-unstable FSFS model sequence is invariant to the level ($n$) of independent replication. Consequently, overall model instabilities may be characterized by those of one observation from the common FSFS model.   
\end{remark}
